---
title: "tree reconstruction"
author: "Richard J Beck"
date: "1/23/2023"
output: html_document
---

Illustrate & test the functions used for analysing karyotype evolution data

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir="~/projects/008_birthrateLandscape/karyotype_evolution/rscripts/")
```

```{r}
source("functions.R")
```


Tests of  the minimal event distance function
```{r}

x1 <- c(2,2,2,2,2,2,2,2)
x2 <- c(4,3,4,4,3,3,2,2)
chr <- c(1,1,2,2,3,3,4,4) 
med(x1,x2,chr) ##expect 4

x1 <- c(2,2,2,2,2,2,2,2)
x2 <- c(2,2,5,5,5,2,2,2)

med(x1,x2,chr)  ##expect 4
med(x2,x1,chr)  ##expect 2


x1 <- c(2,2,2,2,2,2,2,2,2,2)
x2 <- c(4,3,4,4,3,3,2,2,4,4)
chr <- c(1,1,2,2,3,3,4,4,5,5)
med(x1,x2,chr) ##expect 4
med(x2,x1,chr) ##expect 4


```


Generate all intermediate karyotypes between x1 and x2

intermediate karyotypes xi are defined as any karyotype such that
med(x1,xi) + med(xi,x2) == med(x1,x2)

Approach to generate intermediates is:
1) generate all 1MS neighbours xn of x1
2) Set xi = all xn where med(xn,x2) < med(x1,x2)
3) generate all 1MS neighbours xn of xi
4) while (! x2%in% xn) go to step 2

```{r}

x1 <- c(1,2,2,2,2,3)
x2 <- c(4,4,3,3,4,4)
chr <- c(1,1,2,2,3,3)

s1 <- ms_step(x1,x2,chr)
s2 <- ms_step(s1,x2,chr)
s3 <- ms_step(s2,x2,chr)

x1
s1
s2
s3
x2

```

Test the ability to reconstruct evolutionary pathway:

Generate some possible sequence of karyotypes all 1MS apart
delete a karyotype
create MST and infer missing karyotype

```{r}

x <- rbind(c(2,2,2,2,2,2,2,2,2,2),
      c(2,3,2,2,2,2,2,2,2,2),
      c(2,3,2,2,3,3,2,2,2,2),
      c(4,6,4,4,6,6,4,4,4,4),
      c(4,6,3,3,6,6,4,4,4,4),
      c(4,6,3,3,6,6,3,3,4,4),
      c(4,6,3,3,6,6,3,3,4,3))

chr <- c(1,1,2,2,3,3,4,4,5,5)

x <- apply(x,1,paste,collapse=".")
x <- x[c(1,2,4,5,6)]

y <- rep(300,length(x))
names(y) <- x


d <- get_med(y,chr)
dtree <- get_mst(d)

while(max(dtree)>1){
    
    id <- which(dtree>1)[1]
    x1 <- names(y)[id%%nrow(dtree)]
    x2 <- names(y)[ceiling(id/nrow(dtree))]
    
    ibtn <- ms_step(x1,x2,chr,string=T)
    ibtn <- apply(ibtn,1,function(i) paste(i,collapse="."))
    
    x <- rep(0,length(ibtn))
    names(x) <- ibtn
    
    y <- c(y,x)
    
    d <- get_med(y,chr)
    dtree <- get_mst(d)
}

y
```